Exploring the ages of rotating stars using galactic dynamics: a novel approach to calibrating gyrochronology

### 2019-12-12 paper/abstract.tex
small edits

The rotational evolution of cool dwarfs is poorly constrained after $\sim$2-3
billion years due to a lack of precise ages and rotation periods for old
main-sequence stars.
% However, recent evidence suggests that the relationship between rotation
% period and color/temperature/mass does not remain constant over large
% timescales, contradicting a long-held assumption that underlies many
% gyrochronology relations.
In this work, we use the velocities of low-mass \kepler\ dwarfs as an age
proxy, to reveal their rotational evolution and demonstrate that kinematics
could be a useful tool for calibrating gyrochronology.
% At ages less than $\sim$1 Gyr we find that a gyrochronology model, calibrated
% to the Praesepe cluster, accurately predicts the relative ages of stars.
We find that, at young ages, lower-mass stars spin more slowly than
higher-mass stars because their stronger magnetic fields lead to more
efficient angular momentum loss.
This reflects the behavior of young open clusters such as the Pleaides,
Praesepe, and the Hyades.
However, at older ages we find that late G and early K dwarfs rotate at the
same rate as late K dwarfs of the same age, and eventually rotate faster.
% These oThis may be a feature of angular momentum redistribution in the stellar
% interior.
% This implies that low-mass stars lose angular momentum more slowly than
% higher-mass stars after a certain age or Rossby number.
These results align with recent findings from the rotation periods of stars in
middle-aged open clusters and theoretical models that vary the rate of
surface-to-core angular momentum transport as a function of time and mass.
Finally, we find no evidence for mass-dependent heating in a sample of K and M
dwarfs in the \kepler\ field.

>>>>

The rotational evolution of cool dwarfs is poorly constrained after $\sim$2-3
billion years due to a lack of precise ages and rotation periods for old
main-sequence stars.
% However, recent evidence suggests that the relationship between rotation
% period and color/temperature/mass does not remain constant over large
% timescales, contradicting a long-held assumption that underlies many
% gyrochronology relations.
In this work, we use the velocity dispersions of low-mass \kepler\ dwarfs as
an age proxy, to reveal their rotational evolution and demonstrate that
kinematics could be a useful tool for calibrating gyrochronology.
% At ages less than $\sim$1 Gyr we find that a gyrochronology model, calibrated
% to the Praesepe cluster, accurately predicts the relative ages of stars.
We find that the period-\teff\ relationship of the Praesepe cluster does not
apply to stars older than around 1 Gyr.
% : a polynomial gyrochronology model, fit
% to Praesepe stars, under-predicts the ages of late K dwarfs in the field.
Although late-K dwarfs spin more slowly than early-K dwarfs when they are
young, at old ages we find that late-G and early-K dwarfs rotate at the {\it
same rate} or faster than late-K dwarfs of the same age.
% This is probably because their stronger magnetic fields lead to more efficient
% angular momentum loss and reflects the behavior of young open clusters such as
% the Pleaides, Praesepe, and the Hyades.
% However, at older ages we find that late G and early K dwarfs rotate at the
% {\it same rate} or faster than late K dwarfs of the same age.
% These oThis may be a feature of angular momentum redistribution in the stellar
% interior.
% This implies that low-mass stars lose angular momentum more slowly than
% higher-mass stars after a certain age or Rossby number.
This result agrees qualitatively with semi-empirical models that vary the rate
of surface-to-core angular momentum transport as a function of time and mass.
It also aligns with recent observations of stars in the NGC 6811 cluster,
which indicate that K dwarfs experience an epoch of stalled spin-down.
We find that the oldest \kepler\ stars with measured rotation periods are
late-K and early-M dwarfs, indicating that these stars maintain spotted
surfaces and stay magnetically active longer than more massive stars.
Finally, based on their kinematics, we find that many rapidly rotating
K-dwarfs are likely to be synchronized binaries.
% Firstly, we
% find no evidence for mass-dependent heating in a sample of K and M dwarfs in
% the \kepler\ field.

### 2019-12-12 paper/results.tex
small edits

To explore the relationship between rotation period, \teff\ and velocity
dispersion, we first removed high and low velocity outliers from the \mct\
sample by performing 3$\sigma$ sigma-clipping on the \vb\ velocities.
Without sigma-clipping, we found that a small number of high velocity
outliers at the low-temperature end of our sample substantially raised the
velocity dispersion for cooler stars, however the overall trends remain the
same with, or without sigma-clipping.
We also limited the sample to temperatures in the range 5500 K $<$ \teff\ $<$
3500 K to avoid biases caused by the selection function at the faint, cool
end, and binarity or weakened braking \citep{vansaders2016} at the hot end.
Finally, we removed stars with short rotation periods by cutting out stars
with gyrochronal ages less than 0.5 Gyr.

@@

It has been demonstrated that lower-mass stars remain magnetically active
longer than more massive stars, \citep{west2008, kiman2019}.
If the detectability of a rotation period is considered to be a magnetic
activity proxy, then our results provide further evidence for a mass-dependent
activity lifetime.
Figure \ref{fig:vplot} shows that the groups of stars with the largest
velocity dispersions are cooler than 4500 K.
This implies that the oldest stars with detectable rotation periods, are
cooler than 4500 K, \ie\ these-low mass stars stay active longer than more
massive stars.

>>>>

To explore the relationship between rotation period, \teff\ and velocity
dispersion, we first removed high and low velocity outliers from the \mct\
sample by performing 3$\sigma$ sigma-clipping on the \vb\ velocities.
Without sigma-clipping, we found that a small number of high velocity
outliers at the low-temperature end of our sample substantially raised the
velocity dispersion for cooler stars, however the overall trends remain the
same with, or without sigma-clipping.
We also limited the sample to temperatures in the range 5500 K $<$ \teff\ $<$
3500 K to avoid biases caused by the selection function at the faint, cool
end, and binarity or weakened braking \citep{vansaders2016} at the hot end.
Finally, we removed stars with short rotation periods by cutting out stars
with gyrochronal ages less than 0.5 Gyr.
After these cuts, 6879 stars were included in the sample.

@@

It has been demonstrated that lower-mass stars remain magnetically active
longer than more massive stars, \citep[\eg][]{west2008, kiman2019}.
If the detectability of a rotation period is considered to be a magnetic
activity proxy, then our results provide further evidence for a mass-dependent
activity lifetime.
Figure \ref{fig:vplot} shows that the groups of stars with the largest
velocity dispersions are cooler than 4500 K.
This implies that the oldest stars with detectable rotation periods, are
cooler than 4500 K, \ie\ these-low mass stars stay active longer than more
massive stars.
